---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(oomph)
```

# oomph

<!-- badges: start -->
![](https://img.shields.io/badge/cool-useless-green.svg)
[![R-CMD-check](https://github.com/coolbutuseless/oomph/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/coolbutuseless/oomph/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`oomph` is a technical demonstration of using a hash for
fast named lookups of data in lists and vectors.

Internally this is just a hashmap in C which maps strings to integers.  To R, 
this appears as a minimal perfect hash where each string maps to its index.

The hashed lookup can be **1000x** faster than R's standard lookup method (depending on 
number of elements in original object and the number of elements to extract).

Next steps:

* This would be much better with a dynamic hash which could be easily updated
  with new elements.  Currently if even a single name was added/removed or 
  changes, then the entire hash would need to be recalculated.


## Installation

You can install from [GitHub](https://github.com/coolbutuseless/oomph) with:

``` r
# install.package('remotes')
remotes::install_github('coolbutuseless/oomph')
```

## Setup test data


```{r}
library(oomph)
N <- 200000
set.seed(1)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# 200,000 random names
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
nms <- vapply(seq(N), \(i) paste(sample(c(letters, LETTERS, 0:9), 10, T), collapse = ""), character(1))
head(nms)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# A big vector and a big list (each with 200,000 elements)
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
big_vector <- seq(N)
big_list   <- as.list(seq(N))

names(big_vector) <- nms
names(big_list  ) <- nms

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Probe sets to use for testing
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
t1  <- sample(nms,    1, replace = TRUE)
t2  <- sample(nms,   10, replace = TRUE)
t3  <- sample(nms,  100, replace = TRUE)
t4  <- sample(nms, 1000, replace = TRUE)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# By default, the number of hash buckets is the same as the number of 
# strings.  To reduce the possibility of hash collisions (and make lookups
# faster), the number of hash buckets can be changed using the 'size_factor'
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
mph1 <- mph_init(nms)                  # Allocate example length(nms) buckets
mph4 <- mph_init(nms, size_factor = 4) # Allocate: 4 * length(nms) buckets
```


## Compare `match()` with `mph_match()`


```{r match}
bench::mark(
  match(t1, nms),
  mph_match(t1, mph1),
  mph_match(t1, mph4),
  relative = TRUE
)[, 1:5] |> knitr::kable()

bench::mark(
  match(t2, nms),
  mph_match(t2, mph1),
  mph_match(t2, mph4),
  relative = TRUE
)[, 1:5] |> knitr::kable()

bench::mark(
  match(t3, nms),
  mph_match(t3, mph1),
  mph_match(t3, mph4),
  relative = TRUE
)[, 1:5] |> knitr::kable()

bench::mark(
  match(t4, nms),
  mph_match(t4, mph1),
  mph_match(t4, mph4),
  relative = TRUE
)[, 1:5] |> knitr::kable()
```



## List subsetting - Extract 100 elements of a `list` by name

```{r list-subset}
bench::mark(
  `Standard R`           = big_list[t3],
  `[] and mph indexing`  = big_list[mph_match(t3, mph1)],
  `[] and mph indexing`  = big_list[mph_match(t3, mph4)],
  `custom mph method`    = mph_subset(t3, big_list, mph1),
  `custom mph method`    = mph_subset(t3, big_list, mph4),
  check = FALSE
)[, 1:5] |> knitr::kable()
```



## Vector subsetting - Extract 100 elements of a `vector` by name

```{r vector-subset}
bench::mark(
  big_vector[t3],
  big_vector[mph_match(t3, mph1)],
  big_vector[mph_match(t3, mph4)],
  mph_subset(t3, big_vector, mph1),
  mph_subset(t3, big_vector, mph4),
  check = F
)[, 1:5] |> knitr::kable()
```


