---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(oomph)
```

# oomph

<!-- badges: start -->
![](https://img.shields.io/badge/cool-useless-green.svg)
[![R-CMD-check](https://github.com/coolbutuseless/oomph/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/coolbutuseless/oomph/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`oomph` is a packaged for fast named look-up of data in static lists and vectors.

Internally this uses a hash-map in C to map strings to integers.  In R, 
this appears as a minimal perfect hash where each string maps to its index, 
and unknown strings return `NA`

The hashed look-up can be **1000x** faster than R's standard look-up method (depending on 
number of elements in original object and the number of elements to extract).


## What's in the box

* `mph <- mph_init(s, size_factor)` initialise a hash with the given set of strings
    * Using a larger `size_factor` (than the default of `1`) decreases the number 
      of hash collisions, and can make other operations faster at the cost or
      more memory being allocated.
* `mph_match(s, mph)` find the indices of the strings `s` (equivalent to R's `match()`)
* `mph_subset(nms, x, mph)` extract named `elems` from vector/list object `s` using 
  the given hash object.



## Installation

You can install from [GitHub](https://github.com/coolbutuseless/oomph) with:

``` r
# install.package('remotes')
remotes::install_github('coolbutuseless/oomph')
```

## Setup test data


```{r}
library(oomph)
N <- 200000
set.seed(1)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# 200,000 random names
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
nms <- vapply(seq(N), \(i) paste(sample(c(letters, LETTERS, 0:9), 10, T), collapse = ""), character(1))
head(nms)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# A big vector and a big list (each with 200,000 elements)
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
big_vector <- seq(N)
big_list   <- as.list(seq(N))

names(big_vector) <- nms
names(big_list  ) <- nms

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Probe sets to use for testing
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
t1  <- sample(nms,    1, replace = TRUE)
t2  <- sample(nms,   10, replace = TRUE)
t3  <- sample(nms,  100, replace = TRUE)
t4  <- sample(nms, 1000, replace = TRUE)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# By default, the number of hash buckets is the same as the number of 
# strings.  To reduce the possibility of hash collisions (and make look-ups
# faster), the number of hash buckets can be changed using the 'size_factor'
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
mph1 <- mph_init(nms)                  # Allocate example length(nms) buckets
mph4 <- mph_init(nms, size_factor = 4) # Allocate: 4 * length(nms) buckets
```


## Compare `match()` with `mph_match()`


```{r match}
bench::mark(
  match(t1, nms),
  mph_match(t1, mph1),
  mph_match(t1, mph4),
  relative = TRUE
)[, 1:5] |> knitr::kable()

bench::mark(
  match(t2, nms),
  mph_match(t2, mph1),
  mph_match(t2, mph4),
  relative = TRUE
)[, 1:5] |> knitr::kable()

bench::mark(
  match(t3, nms),
  mph_match(t3, mph1),
  mph_match(t3, mph4),
  relative = TRUE
)[, 1:5] |> knitr::kable()

bench::mark(
  match(t4, nms),
  mph_match(t4, mph1),
  mph_match(t4, mph4),
  relative = TRUE
)[, 1:5] |> knitr::kable()
```



## List subsetting - Extract 100 elements of a `list` by name

```{r list-subset}
bench::mark(
  `Standard R`           = big_list[t3],
  `[] and mph indexing`  = big_list[mph_match(t3, mph1)],
  `[] and mph indexing`  = big_list[mph_match(t3, mph4)],
  `custom mph method`    = mph_subset(t3, big_list, mph1),
  `custom mph method`    = mph_subset(t3, big_list, mph4),
  check = FALSE
)[, 1:5] |> knitr::kable()
```



## Vector subsetting - Extract 100 elements of a `vector` by name

```{r vector-subset}
bench::mark(
  big_vector[t3],
  big_vector[mph_match(t3, mph1)],
  big_vector[mph_match(t3, mph4)],
  mph_subset(t3, big_vector, mph1),
  mph_subset(t3, big_vector, mph4),
  check = F
)[, 1:5] |> knitr::kable()
```



## Comparison to R's hashed look-ups in environments

```{r compare-env}
ee <- as.environment(big_list)

bench::mark(
  `Standard R`           = big_list[t3],
  `[] and mph indexing`  = big_list[mph_match(t3, mph1)],
  `[] and mph indexing`  = big_list[mph_match(t3, mph4)],
  `custom mph method`    = mph_subset(t3, big_list, mph1),
  `custom mph method`    = mph_subset(t3, big_list, mph4),
  `R hashed environment` = mget(t3, ee),
  check = FALSE
)[, 1:5] |> knitr::kable()
```

## Time taken to build the hash

```{r}
set.seed(1)
N <- 1000
nms1k <- vapply(seq(N), \(i) paste(sample(c(letters, LETTERS, 0:9), 10, T), collapse = ""), character(1))

N <- 10000
nms10k <- vapply(seq(N), \(i) paste(sample(c(letters, LETTERS, 0:9), 10, T), collapse = ""), character(1))

N <- 100000
nms100k <- vapply(seq(N), \(i) paste(sample(c(letters, LETTERS, 0:9), 10, T), collapse = ""), character(1))

bench::mark(
  mph_init(nms1k),
  mph_init(nms10k),
  mph_init(nms100k),
  check = FALSE
)[, 1:5] |> knitr::kable()

```






