---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(oomph)
```

# oomph

<!-- badges: start -->
![](https://img.shields.io/badge/cool-useless-green.svg)
[![R-CMD-check](https://github.com/coolbutuseless/oomph-dev/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/coolbutuseless/oomph-dev/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`oomph` is a technical demonstration of using a *minimal perfect hashing function* to do
fast named lookups of data in lists and vectors.

The hashed lookup can be **1000x** faster than R's standard lookup method (depending on 
number of elements to extract).

Next steps:

* This would be much better with a dynamic minimal perfect hash which updated
  with new elements
* The included mph code has some issues with large numbers of input strings.
  I think it may have something to do with large stack using with the
  recursive calling of `graph.c / cyclicdeledge()`


## Installation

You can install from [GitHub](https://github.com/coolbutuseless/oomph) with:

``` r
# install.package('remotes')
remotes::install_github('coolbutuseless/oomph')
```

## Setup test data


```{r}
library(oomph)
N <- 200000
set.seed(1)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# 200,000 random names
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
nms <- vapply(seq(N), \(i) paste(sample(c(letters, LETTERS, 0:9), 10, T), collapse = ""), character(1))
head(nms)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# A big vector and a big list (each with 200,000 elements)
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
big_vector <- seq(N)
big_list   <- as.list(seq(N))

names(big_vector) <- nms
names(big_list  ) <- nms

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Probe sets to use for testing
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
t1  <- sample(nms,    1, replace = TRUE)
t2  <- sample(nms,   10, replace = TRUE)
t3  <- sample(nms,  100, replace = TRUE)
t4  <- sample(nms, 1000, replace = TRUE)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# 
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
mph <- mph_init(nms)
```


## Compare `match()` with `mph_match()`


```{r match}
bench::mark(
  match(t1, nms),
  mph_match(t1, mph),
  relative = TRUE
)[, 1:5] |> knitr::kable()

bench::mark(
  match(t2, nms),
  mph_match(t2, mph),
  relative = TRUE
)[, 1:5] |> knitr::kable()

bench::mark(
  match(t3, nms),
  mph_match(t3, mph),
  relative = TRUE
)[, 1:5] |> knitr::kable()

bench::mark(
  match(t4, nms),
  mph_match(t4, mph),
  relative = TRUE
)[, 1:5] |> knitr::kable()
```



## List subsetting - Extract 100 elements of a `list` by name

```{r list-subset}
bench::mark(
  `Standard R`           = big_list[t3],
  `[] and mph indexing`  = big_list[mph_match(t3, mph)],
  `custom mph method`    = mph_subset(big_list, t3, mph),
  check = FALSE
)[, 1:5] |> knitr::kable()
```



## Vector subsetting - Extract 100 elements of a `vector` by name

```{r vector-subset}
bench::mark(
  big_vector[t3],
  big_vector[mph_match(t3, mph)],
  mph_subset(big_vector, t3, mph),
  check = F
)[, 1:5] |> knitr::kable()
```



## Creating the C code for the minimal perfect hash if using 'mph' on the command line.

There are a number of programs which will produce C code to
do minimal perfect hashing of a list of strings e.g. `gperf`, `cmph`.

I chose `mph` as described in [A Family of Perfect Hashing Methods(pdf)](https://staff.itee.uq.edu.au/havas/TR0242.pdf)

* Download and compile [http://www.ibiblio.org/pub/Linux/devel/lang/c/mph-1.2.tar.gz]
* In R: `writeLines(colors(), "input.txt")`
* On the command line:  `mph < input.txt | emitc > hash.c`
* Incorporate the `hash.c` code into this package and add the `col_to_rgb()` wrapper
  which uses it.



