---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(oomph)
```

# oomph

<!-- badges: start -->
![](https://img.shields.io/badge/cool-useless-green.svg)
[![R-CMD-check](https://github.com/coolbutuseless/oomph/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/coolbutuseless/oomph/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`oomph` is a package for fast string matching within a static set of strings - 
where each string maps to its integer position.

This is useful for fast named look-up in fixed lists and vectors.

Internally this uses a hashmap written in C to map strings to integers.  In R, 
this appears as a minimal perfect hash where each string maps to its position in the 
original data (unknown strings will return `NA`).

The hashed look-up can be more than **1000x** faster than R's standard look-up method (depending on 
number of elements in original object and the number of elements to extract).


## What's in the box

* `mph <- mph_init(s)` initialise a hashmap with the given set of strings
* `mph_match(s, mph)` find the indices of the strings `s` (equivalent to R's `match()`)



## Installation

You can install from [GitHub](https://github.com/coolbutuseless/oomph) with:

``` r
# install.package('remotes')
remotes::install_github('coolbutuseless/oomph')
```

## Simple example

```{r}
mph <- mph_init(c('mary', 'had', 'a', 'little', 'lamb'))
mph_match('mary', mph)
mph_match('lamb', mph)
mph_match('monkey', mph) # not a string from the original vector.
```


## Benchmarking

### Setup benchmark data


```{r}
library(oomph)
N <- 500000
set.seed(1)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# 500k random names
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
nms <- vapply(seq(N), \(i) paste(sample(c(letters, LETTERS, 0:9), 10, T), collapse = ""), character(1))
head(nms)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# A big named vector and named list (each with 500k elements)
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
big_vector <- seq(N)
big_list   <- as.list(seq(N))

names(big_vector) <- nms
names(big_list  ) <- nms

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Probe sets to use for testing
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
t2  <- sample(nms,   100, replace = TRUE)
t3  <- sample(nms,  1000, replace = TRUE)
t4  <- sample(nms, 10000, replace = TRUE)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# By default, the number of hashmap buckets is twice the number of 
# strings.  To reduce the possibility of hash collisions (and possibly make look-ups
# faster), the number of hashmap buckets can be increased by increasing 'size_factor'
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
mph <- mph_init(nms) 
```


### Compare base R `match()` with oomph's `mph_match()`


```{r match}
bench::mark(
  match(t2, nms),
  mph_match(t2, mph),
  check = TRUE
)[, 1:5] |> knitr::kable()

bench::mark(
  match(t3, nms),
  mph_match(t3, mph),
  check = TRUE
)[, 1:5] |> knitr::kable()

bench::mark(
  match(t4, nms),
  mph_match(t4, mph),
  check = TRUE
)[, 1:5] |> knitr::kable()
```




### Vector subsetting - Extract 100 elements of a `vector` by name

```{r vector-subset}
bench::mark(
  big_vector[t2],
  big_vector[mph_match(t2, mph)],
  check = TRUE
)[, 1:5] |> knitr::kable()
```


### List subsetting - Extract 100 elements of a `list` by name

Also compare to using hashed named lookup in a standard R environment

```{r list-subset}
ee <- as.environment(big_list)

bench::mark(
  `Standard R`           = big_list[t2],
  `R hashed environment` = mget(t2, ee),
  `[] and mph indexing`  = big_list[mph_match(t2, mph)],
  check = TRUE
)[, 1:5] |> knitr::kable()
```



### Time taken to build the hashmap

```{r}
set.seed(1)
chrs <- c(letters, LETTERS, 0:9)
N <- 1000
nms1k <- vapply(seq(N), \(i) paste(sample(chrs, 10, T), collapse = ""), character(1))

N <- 10000
nms10k <- vapply(seq(N), \(i) paste(sample(chrs, 10, T), collapse = ""), character(1))

N <- 100000
nms100k <- vapply(seq(N), \(i) paste(sample(chrs, 10, T), collapse = ""), character(1))

bench::mark(
  mph_init(nms1k),
  mph_init(nms10k),
  mph_init(nms100k),
  check = TRUE
)[, 1:5] |> knitr::kable()

```

